---
title: "Clustering Example 1: Five Data Points"
author: "Amy Wagaman, Nathan Carter"
date: "July 2020"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

## Load necessary library.

```{r message=FALSE, warning=FALSE}
library(ggformula)
```

## Generate 5 data points and graph them.

```{r}
x <- c(4,2,4.5,3,1)
y <- c(5.5,2,6,7,3)
Label <- c("(4, 5.5)", "(2, 2)", "(4.5, 6)", "(3, 7)", "(1, 3)")
mydata <- data.frame(x, y, Label)
ggplot(mydata,
       aes(x=x, y=y, label=Label))+
       xlim(0.75, 4.75)+ylim(1.5, 7.5)+geom_point()+
       geom_text(aes(label=Label),hjust = 0.5,vjust = -0.75)
```

## Convert to a data frame and compute the distance matrix.

```{r}
mydata <- data.frame(x,y)
mydata
round(dist(mydata),2)
```

## Perform hierarchical clustering with single linkage.

```{r}
plot(hclust(dist(mydata), method = "single"))
```

## Perform hierarchical clustering with complete linkage.

The clusters are the same as in the previous case, but there are minor changes to the heights of the bars in the dendrogram.

```{r}
plot(hclust(dist(mydata), method = "complete"))
```

## Perform hierarchical clustering with average linkage.

```{r}
plot(hclust(dist(mydata), method = "average"))
```

## Perform hierarchical clustering with Ward's method.

```{r}
plot(hclust(dist(mydata), method = "ward.D"))
```
