---
title: "Clustering Example 3: $k$-means Applied to Wine Data"
author: "Amy Wagaman, Nathan Carter"
date: "July 2020"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

## Load necessary libraries.

```{r message=FALSE, warning=FALSE}
library(mosaic)
library(cluster)
```

## Load wine dataset.

This requires you to have access to the `winedata.txt` file.  It is available in the book's GitHub repository at the following URL.

[https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/winedata.txt](https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/winedata.txt)

If you run this R code, place the data file in the same folder as the code file.

```{r}
winedata <- read.csv("winedata.txt")
```

## See column names, data types, and some values in each column.

```{r}
glimpse(winedata)
```

## See summary statistics for all columns in wine data.

```{r}
favstats(~ Alcohol, data = winedata)
favstats(~ MalicAcid, data = winedata)
favstats(~ Ash, data = winedata)
favstats(~ AlcalinityofAsh, data = winedata)
favstats(~ Magnesium, data = winedata)
favstats(~ TotalPhenols, data = winedata)
favstats(~ Flavanoids, data = winedata)
favstats(~ NonflavanoidPhenols, data = winedata)
favstats(~ Proanthocyanins, data = winedata)
favstats(~ ColorIntensity, data = winedata)
favstats(~ Hue, data = winedata)
favstats(~ ODDilutedWines, data = winedata)
favstats(~ Proline, data = winedata)
```

## Run $k$-means clustering for $k=1$ to $k=12$.

We plot the within-groups sum of squares for each run, so that we can assess which $k$ value may be best.  Note the choice of a random number seed, for reproducibility.

```{r}
set.seed(240)
wss <- rep(0, 12) #creates 12 copies of 0 to create an empty vector
for(i in 1:12) {
  wss[i] <- sum(kmeans(scale(winedata[, -c(1)]), centers = i)$withinss)
}
plot(1:12, wss, type = "b",
     xlab = "Number of clusters", ylab = "Within groups sum of squares")
```

## Run $k$-means with the chosen value of $k=3$.

And print the results, which include cluster centroids, the clustering partition as a vector, and the within-groups sum of squares.

```{r}
set.seed(304)
Ksol1 <- kmeans(scale(winedata[, -c(1)]), centers = 3) #centers is the # of clusters
list(Ksol1) #so you can see what it gives you
```

## Create corresponding silhouette plot.

```{r}
kmeansSil <- silhouette(Ksol1$cluster, dist(scale(winedata[, -c(1)])))
silsum <- summary(kmeansSil)
plot(kmeansSil, col = "black")
```

## Report average silhouette width more precisely.

```{r}
summary(kmeansSil)$avg.width
```

## Compute average silhouette width for all possible values of $k$.

Thus $k$ ranges from 2 to $n-1$ (where $n$ is the number of observations, here 178) and we report summary statistics for the collection of silhouette widths.

```{r}
n <- 178
mydist <- dist(scale(winedata[, -c(1)]))
avgwidths <- rep(0, 176)

for (i in 2:(n-1)) {
  Ksol <- kmeans(scale(winedata[, -c(1)]), centers = i) #centers is the # of clusters
  kmeansSil <- silhouette(Ksol$cluster, mydist)
  avgwidths[i-1] <- summary(kmeansSil)$avg.width
}

summary(avgwidths)
```

## Plot the $k=3$ solution in principal component space.

```{r}
set.seed(1)
winePCAs <- princomp(winedata[, -c(1)], cor = TRUE)
plot(winePCAs$scores[, 1:2], type = "n",
     xlab = "Principal Component 1", ylab = "Principal Component 2",
     main="K-means Three-Cluster Solution")
text(winePCAs$scores[, 1:2], labels = Ksol1$cluster, cex = 0.7)
```

## Compare found clusters to original wine cultivars.

```{r}
tally(winedata$Cultivar ~ Ksol1$cluster)
```

## Perform $k$-means clustering with $k=4$.

```{r}
set.seed(304)
Ksol2 <- kmeans(scale(winedata[, -c(1)]), centers = 4) #centers is the # of clusters
list(Ksol2)
```

## Compare $k=4$ solution to original wine cultivars.

```{r}
tally(Ksol1$cluster ~ Ksol2$cluster)
```


