---
title: "Clustering Example 4: Hierarchical Clustering of Wine Data"
author: "Amy Wagaman, Nathan Carter"
date: "July 2020"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

## Load necessary libraries.

```{r message=FALSE, warning=FALSE}
library(mosaic)
library(cluster)
```

This requires you to have access to the `winedata.txt` file.  It is available in the book's GitHub repository at the following URL.

[https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/winedata.txt](https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/winedata.txt)

If you run this R code, place the data file in the same folder as the code file.

```{r}
winedata <- read.csv("winedata.txt")
```

For summaries of the variables in the wine data, see the Clustering Example 3 file in the same folder as this one.

## Compute a distance matrix for the wine data.

```{r}
wine.dist.scale <- dist(scale(winedata[, -c(1)]))
```

## Apply hierarchical clustering with single linkage.

```{r}
hcsingle <- hclust(wine.dist.scale, method = "single") 
list(hcsingle) # reminds you of properties of the solution, if desired
```

## Plot the resulting dendrogram.

```{r}
plot(hcsingle, cex = 0.7)
```

## Extract a single clusering solution from the dendrogram.

Here we specify the height $h$ that determines the clustering.  We display just the number of observations in each cluster.

```{r}
singleSol <- (cutree(hcsingle, h = 2.25)) #cluster labels are numeric, h= height
summary(as.factor(singleSol)) #as factor to get table
```

## Apply hierarchical clustering with complete linkage and plot dendrogram.

```{r}
hccomp <- hclust(wine.dist.scale, method = "complete") 
plot(hccomp, cex = 0.7)
```

## Extract a single clusering solution from the dendrogram.

Here we specify the number $k$ of clusters that we desire in the solution.  We display just the number of observations in each cluster.

```{r}
compSol <- (cutree(hccomp, k= 4)) #k= # clusters, k=3 also reasonable
summary(as.factor(compSol)) #as factor to get table
```

## Apply hierarchical clustering with Ward's method and plot dendrogram.

```{r}
hcward <- hclust(wine.dist.scale, method = "ward.D") 
plot(hcward, cex = 0.7)
```

## Extract a single clusering solution from the dendrogram.

Here we specify the number $k$ of clusters that we desire in the solution.  We display just the number of observations in each cluster.

```{r}
wardSol <- (cutree(hcward, k= 3)) #k= # clusters, k=3 also reasonable
summary(as.factor(wardSol)) #as factor to get table
```
