---
title: "Clustering Example 5: Model-based Clustering of Wine Data"
author: "Amy Wagaman, Nathan Carter"
date: "July 2020"
output: 
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---

## Load necessary libraries.

```{r message=FALSE, warning=FALSE}
library(mosaic)
library(mclust)
library(readr)
```

This requires you to have access to the `winedata.txt` file.  It is available in the book's GitHub repository at the following URL.

[https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/winedata.txt](https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/winedata.txt)

If you run this R code, place the data file in the same folder as the code file.

```{r}
winedata <- read.csv("winedata.txt")
```

For summaries of the variables in the wine data, see the Clustering Example 3 file in the same folder as this one.

## Plot BIC for various model-based methods.

```{r}
mclustsol <- mclustBIC(winedata[, -1])
plot(mclustsol)
```

## Compare two mixture models for a small range of $k$ values.

This requires you to have access to the `mclustBICs.csv` file.  It is available in the book's GitHub repository at the following URL.

[https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/mclustBICs.csv](https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/mclustBICs.csv)

If you run this R code, place the data file in the same folder as the code file.

```{r}
mclustBICs <- read_csv("mclustBICs.csv")
mclustBICs2 <- mutate(mclustBICs, Model = factor(Model))
mclustBICs3 <- filter(mclustBICs2, Model == "EVE" | Model == "VVE", k < 6, k > 1.5)

gf_point(BIC ~ k, data = mclustBICs3, shape = ~ Model) %>%
  gf_labs(title = "BIC Comparison of EVE and VVE Mixture Models",
          x = "Number of Components") + geom_line()
summary(mclustsol)
```

## Fit Gaussian finite mixture models to the data.

```{r}
mod1 <- Mclust(winedata[, -1], x = mclustsol)
summary(mod1, parameters = TRUE)
```

## Plotting with so many variables does not lead to readable results.

```{r}
plot(mod1, what = "classification")
```

## Plot the clustering from the VVE model with $k=3$, using principal component space.

```{r}
set.seed(1)
winePCAs <- princomp(winedata[, -c(1)], cor = TRUE)
plot(winePCAs$scores[, 1:2], type = "n", 
     xlab = "Principal Component 1", ylab = "Principal Component 2", 
     main="MClust VVE Three-Cluster Solution")
text(winePCAs$scores[, 1:2], labels = mod1$classification, cex = 0.7)
```

## Compare that solution to the original wine cultivars.

```{r}
tally(winedata$Cultivar ~ mod1$classification)
```

