---
title: "Clustering Example 7: Network Clustering of Dolphin Data"
author: "Amy Wagaman, Nathan Carter"
date: "July 2020"
output: 
  pdf_document:
    fig_height: 5
    fig_width: 5
  html_document:
    fig_height: 5
    fig_width: 5
  word_document:
    fig_height: 5
    fig_width: 5
---

## Load necessary libraries.

```{r message=FALSE, warning=FALSE}
library(igraph)
library(ape)
```

This requires you to have access to the `dolphins.gml` file.  It is available in the book's GitHub repository at the following URL.

[https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/dolphins.gml](https://github.com/ds4m/ds4m.github.io/tree/master/chapter-5-resources/dolphins.gml)

If you run this R code, place the data file in the same folder as the code file.

How many vertices and edges does the graph have?

```{r}
g <- read_graph("dolphins.gml", format = "gml")
vcount(g)
ecount(g)
```

## Plot the graph

Some layout algorithms use randomness, so we set a seed, for reproducibility.

```{r}
set.seed(540)
mylayout3 <- layout_(g, with_gem())
igraph.options(vertex.size=5, vertex.color = "NA", vertex.shape = "circle",
               vertex.label.cex = 0.5, vertex.label.dist = 1.3)
plot(g, layout = mylayout3)
```

## We can ask questions about network density.

```{r}
degree(g)
summary(degree(g))
```

## Apply the `fastgreedy.community` clustering algorithm.

```{r}
set.seed(240)
fg1 <- fastgreedy.community(g)
length(fg1) # How many clusters?
sizes(fg1) # How large was each cluster?
membership(fg1) # the familiar clustering vector
```

## Plot the network again, with clusters shown.

```{r}
plot(fg1, g)
```

## Measure the clustering quality.

The `fastgreedy.community` algorithm seeks to maximize modularity, which is a measure of how separated different vertex types are from one another.  We can ask for the modularity of the result.

```{r}
modularity(g, fg1$membership)
```

## Plot the corresponding dendrogram.

Because the `fastgreedy.community` algorithm is hierarchical, it also produces a dendrogram.

```{r}
dendPlot(fg1, mode="phylo", palette = gray(c(0, 0.5, 0, 0.5)),
         cex = 0.5, label.offset = 0.1)
```

## Improve the clustering plot.

The plot that moves vertices near to one another is a bit busy and messy.  Let's assign vertex shapes to indicate the clustering solution and plot without the colored blobs shown above.

```{r}
shapes <- rep("circle", 62)
colors <- rep("NA", 62)

for (i in 1:62) {
  if (fg1$membership[i]==2) { shapes[i] = "square" }
  if (fg1$membership[i]==4) { shapes[i] = "square" }
  if (fg1$membership[i]==1) { colors[i] = "gray"   }
  if (fg1$membership[i]==2) { colors[i] = "gray"   }
}

igraph.options(vertex.size=5, vertex.color = colors, vertex.label.cex = 0.5,
               vertex.label.dist = 1.3, vertex.shape = shapes)
plot(g, layout = mylayout3)
```

## Apply the `leading.eigenvector.community` clustering algorithm.

```{r}
sp1 <- leading.eigenvector.community(g)
length(sp1) # How many clusters?
sizes(sp1) # How large was each cluster?
membership(sp1) # the familiar clustering vector
```

## Plot the network again, with clusters shown.

```{r}
plot(sp1, g)
```

## Measure the clustering quality.

See comments above about the `modularity` metric.

```{r}
modularity(g, sp1$membership)
```

## Apply the `walktrap.community` clustering algorithm.

```{r}
wt1 <- walktrap.community(g)
length(wt1) # How many clusters?
sizes(wt1) # How large was each cluster?
membership(wt1) # the familiar clustering vector
```

## Plot the network again, with clusters shown.

```{r}
plot(wt1, g)
```

## Measure the clustering quality.

See comments above about the `modularity` metric.

```{r}
modularity(g, wt1$membership)
```
